Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-04-04 draft.tex
Edited the introduction.

A nearly overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the vast majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in many contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Need to add some more recent references here.} 
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,
slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}.
In this paper, we revisit and update astrophysical constraints on SIDM models from 
supernova cooling, finding constraints that are considerably less restrictive than in 
previous work. 
\arz{Might want to add a comment about SIDM affecting small-scale structure in CDM, the satellites problems and core/cusp problems.}

SIDM models in which large self-interaction cross sections are mediated by sufficiently light 
bosons ($M \lesssim 100$~GeV) can be constrained astrophysically using supernovae, particularly 
SN1987A. Light gauge bosons will be produced within the hot supernovae core and 
radiated from the supernova. This non-standard energy loss mechanism can result in an energy loss rate 
from the supernova core that is inconsistent with observations of SN1987A, analogous to the 
classic constraint on axions \cite{turner88,raffelt96_book}. This mechanism was already exploited by 
\cite{dent_etal12} (and \cite{rrapaj_reddy16}, with a slight modification) (\arz{Cameron, add others of which you may be aware.}) 
to constraint dark electromagnetism models of SIDM in which the self-interaction arises from a Yukawa potential 
and the vector boson is kinetically mixed with the standard model photon.

As we mentioned above, we find SN1987A constraints on dark photon SIDM models that are considerably different 
from those of \cite{dent_etal12}. If correct, our result has important consequences. 
The Snowmass white paper by Kaplinghat, Tulin, and Yu \cite{kaplinghat_etal13_whitepaper} nicely 
summarizes a variety of constraints on dark electromagnetism models of SIDM (many details can be 
found in the the very informative work of Ref.~\cite{bjorken_etal09}). 
One point that is clear from Ref.~\cite{kaplinghat_etal13_whitepaper} is that there exists 
only a narrow range of viable photon mixing parameters that can lead to production of SIDM in the early universe and 
interesting effects on cosmological structure formation, while simultaneously evading all constraints, 
including the constraint of Ref.~\cite{dent_etal12}. Our results suggest that the SN1987A constraints from 
Ref.~\cite{dent_etal12} are too restrictive. The constraints that we quote reveal a region of unrestricted mixing 
parameters that ranges over two orders of magnitude. Interestingly, the parameter space opened by our updated 
constraints corresponds to SIDM models that may have a significant impact on cosmological structure formation, 
particularly the structures of dark matter halos and galaxies.

The discrepancy between our result and that of Ref.~\cite{dent_etal12} arises due to multiple factors. First, it is straightforward to 
demonstrate that the kinematical relationships in given in Appendix A of Ref.~\cite{dent_etal12} are incorrect. Second, the 
squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} are incorrect. Our squared matrix elements 
are considerably more complicated. Multiple errors must have led to the incorrect squared matrix elements in 
Ref.~\cite{dent_etal12}. At the very least, these errors include simplification using the incorrect kinematics and neglect of the 
mass of the gauge boson (which, while legitimate for the $\sim$meV-mass axion, is not justified in this context). 
Discrepancies in addition to those kinematical issues must also be present, 
the causes of which are not apparent from the exposition, because the squared matrix 
elements in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} do not obey the 
correct symmetries under interchange of the incoming or outgoing baryon momenta. 
Given these errors, we suggest that our results should supercede these previously-published constraints.

The remainder of this paper is organized as follows. In Section~\ref{section:model}, we discuss 
dark photon models. We describe our calculation of SN1987A constraints on SIDM in 
Section~\ref{section:computational} and present our primary results in Section~\ref{section:results}. 
We summarize our work and draw conclusions in Section~\ref{section:conclusions}.

@@

%----- Computational Details
\section{Methods}
\label{section:computational}

@@

The brehmsstrahlung process is not the only process with which we must be concerned. Clearly, the rate of 
brehmsstrahlung will increase with $\varepsilon$; however, $\varepsilon$ can become sufficiently large 
that the radiated gauge bosons do not escape the supernova. This happens if the $\Aprime$ either decay to 
standard model particles or interact with standard model particles prior to exiting the supernova core. 
In either case, the energy is not lost and the $\Aprime$ does not provide a cooling channel for the supernova. 
Consequently, for a given $m_{\Aprime}$, there is a maximum $\varepsilon$ that can be constrained in this 
manner. We estimate $\Aprime$ decay and scattering probabilities, and the upper limits on the $\varepsilon$ 
constraints in this section as well.

\subsection{Brehmsstrahlung of $\Aprime$ Bosons}

@@

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\varepsilon \approx  1e-8-9 $.
\arz{Cameron, some similar comments as for Figure 1. Can we write numbers as "$10^{-9}$" rather 
than "1e-09" and so on? Can we make the axis labels and tick mark labels larger?}}
\label{fig:constraint}
\end{figure}

\begin{figure}[th]
\includegraphics[width=9cm]{stages.eps}
\caption{The changes in the bounds generated as each stage of the calculation departs from Ref.~\cite{dent_etal12}. Interestingly the kinematics have apparently relatively little effect.
\arz{Cameron, several to-do items for this figure. On the y-axis, the symbol should be "$\varepsilon$" rather than 
"e". Can we write numbers as "$10^{-9}$" rather than "1e-09" and so on? Can we make the axis labels and 
tick mark labels larger? Can you make all of the lines thicker? Can you make each of the four lines a distinct color?}
}
\end{figure}

>>>>

An overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in numerous contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Cameron: Please review the literature and add some more recent references here. 
I know that Annika Peter has a few papers and Jonathan Feng probably has some more. 
Ensuring that you are up to date on these papers is important.} 
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,
slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}. 
In this paper, we revisit and update astrophysical constraints on SIDM models from 
supernova cooling.

SIDM models in which large self-interaction cross sections are mediated by sufficiently light 
bosons ($M \lesssim 100$~GeV) can be constrained astrophysically using supernovae, particularly 
SN1987A. Light gauge bosons will be produced within the hot supernova core, primarily through 
brehmsstrahlung, and radiated. This non-standard energy loss mechanism can result in an energy loss rate 
from the supernova core that is inconsistent with observations of neutrinos from SN1987A, 
analogous to the classic constraint on axions \cite{turner88,raffelt96_book}. 
This effect has already been exploited by Dent et al. \cite{dent_etal12}, Rrapaj and Reddy \cite{rrapaj_reddy16}, and 
Hardy and Lasenby \cite{hardy_lasenby17} \arz{Cameron, add others of which you may be aware.}
to constraint dark electromagnetism models in which the new gauge boson, the so-called dark photon, 
is kinetically mixed with the standard model photon. The SN1987A constraint places a limit on the mixing parameter.

We initiated our study because of a number of ambiguities appearing in the previous literature on this subject. 
In particular, we could not reproduce the constraints of Ref.~\cite{dent_etal12}. During the course of our 
study, we identified a number of errors in the analysis of Ref.~\cite{dent_etal12}. First, it is straightforward 
to demonstrate that the kinematical relationships given in Appendix A of Ref.~\cite{dent_etal12} are 
incorrect. Second, the squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} must 
be incorrect. We have not been able to identify the errors which led to these incorrect expressions; however, 
there must have been several oversights in the analysis of Ref.~\cite{dent_etal12}. For example, the squared 
matrix elements in Eq.~(A3) and Eq.~(C18) do not obey the correct symmetries under interchange of incoming 
and/or outgoing momenta. Furthermore, Ref.~\cite{dent_etal12} neglects the mass of the gauge boson, which 
is legitimate in the classic case of the $\sim$meV-mass axion, but not in the present context. Finally, Ref.~\cite{dent_etal12} 
employed an inconsistent model for the permitted energy loss rate from the supernova interior. Our work amounts 
primarily to repeating the calculation of Ref.~\cite{dent_etal12} in order to rectify these oversights. Our primary 
calculation treats nucleon scattering via one-pion exchange (OPE). We have repeated this analysis 
using and effective field theory for the nucleon interaction and find very similar results so we do not 
report on those in detail. \arz{I think we may want to add back in the EFT calculation for this paper. 
That would at least give this paper something a bit unique when compared to the other papers on 
the subject.}

As we were completing our manuscript, two related papers were published. Rrapaj and Reddy \cite{rrapaj_reddy16} 
computed bounds on mixing of the dark and standard model photons using a soft radiation approximation (SRA) 
for dark photon Brehmsstrahlung. This has the distinct advantage of enabling nucleon scattering data to be used 
directly in the estimation of the Brehmsstrahlung rate, but is only approximate because at large dark photon masses 
the radiated dark photons carry off considerable momentum and energy. We have been able to reproduce the 
result of Rrapaj and Reddy \cite{rrapaj_reddy16} and find that the SRA plausibly results in only a factor or 
$\sim 3$ underestimate of the upper bound on the dark photon mixing parameter at most. Moreover, 
our OPE results agree well with the SRA calculation of Rrapaj and Reddy \cite{rrapaj_reddy16}. Nonetheless, 
Ref.~\cite{rrapaj_reddy16} quote results very similar to those of Dent et al. \cite{dent_etal12}. We find, 
rather remarkably, that the various errors in the analysis of Ref.~\cite{dent_etal12} conspire to yield a constraint that 
is very similar to the correct answer.

More recently, Hardy and Lasenby \cite{hardy_lasenby17} studied bounds on these same models (and others) 
including plasma effects. For simplicity, we have not included these plasma effects in our calculations. Hardy 
and Lasenby base their calculation off of the SRA of Ref.~\cite{rrapaj_reddy16}. Consequently, 
for dark photon masses $\gtrsim 10$~MeV they find very similar results to Ref.~\cite{rrapaj_reddy16} as well 
as the constraints we present in this manuscript. For dark photon masses $\lesssim 10$~MeV, Hardy and 
Lasenby demonstrate that constraints on the dark photon mixing parameter are significantly weaker than 
one finds when neglecting plasma effects \cite{hardy_lasenby17}.

It is important to delineate correctly the range of the viable parameter space for SIDM models. 
The parameter space available to dark electromagnetism models of SIDM has been studied extensively 
not only in the aforementioned papers, but also in the work of Bjorken et al. \cite{bjorken_etal09} and 
the Snowmass white paper by Kaplinghat, Tulin, and Yu \cite{kaplinghat_etal13_whitepaper}. One 
point that is clear from previous work is that there is at most only a slim sliver of parameter space that 
can simultaneously yield the correct relic abundance of SIDM through thermal production, 
have interesting effects on cosmological structure formation, and 
evade all constraints including the constraints from SN1987A. 
In accord with Ref.~\cite{rrapaj_reddy16} and Ref.~\cite{hardy_lasenby17}, we find that 
the constraints quoted in Ref.~\cite{dent_etal12} are too restrictive, but only by a factor 
of $\sim 4$ due to a conspiratorial cancellation of errors in Ref.~\cite{dent_etal12}. 
Hardy and Lasenby go on to demonstrate that these constraints are significantly too 
restrictive for dark photon masses $\lesssim 10$~MeV due to plasma effects.

The remainder of this paper is organized as follows. In Section~\ref{section:model}, we discuss 
dark photon models. We describe our calculation of SN1987A constraints on SIDM in 
Section~\ref{section:computational} and present our primary results in Section~\ref{section:results}. 
We stress only those points that are key to understanding the relationship between our work 
and the work of both Dent et al. in Ref.~\cite{dent_etal12} and Rrapaj and Reddy in Ref.~\cite{rrapaj_reddy16}. 
We summarize our work and draw conclusions in Section~\ref{section:conclusions}.

@@

%----- Methods
\section{Methods}
\label{section:computational}

@@

The brehmsstrahlung process is not the only process with which we must be concerned. Clearly, the rate of 
brehmsstrahlung will increase with $\varepsilon$; however, $\varepsilon$ can become sufficiently large 
that the radiated gauge bosons do not escape the supernova. This happens if the $\Aprime$ particles 
either decay to or interact with standard model particles prior to exiting the supernova core. 
In either case, the energy is not lost and the $\Aprime$ does not provide a cooling channel for the supernova. 
Consequently, for a given $m_{\Aprime}$, there is a maximum $\varepsilon$ that can be constrained in this 
manner. We estimate $\Aprime$ decay and scattering probabilities, and the upper limits on the $\varepsilon$ 
constraints in this section as well. However, we note that terrestrial experiments generally rule out 
mixing parameters higher than the upper limits of the SN1987A forbidden region, so a precise estimate 
of these upper limits is not necessary.

%-- Brehmsstrahlung--	
\subsection{Brehmsstrahlung of $\Aprime$ Bosons}

@@

\begin{figure*}[th]
\includegraphics[width=14cm]{constraint.pdf}
\caption{
Constraints on dark photon models. 
}
\label{fig:constraints}
\end{figure*}
